Drug sensitivity prediction with the normal inverse Gaussian prior informed by external data

### 2019-11-22 docs/manuscript1.Rnw
commit before rerunning analysis_gdsc.R

	We estimate four models: (i) an NIG model that does include external 
	covariates in the estimation, (ii) the NIG model estimated as in Section 
	\ref{sec:estimation}, and regular (iii) lasso and (iv) ridge models.
	Exclusion of the external covariates as in model (i)
	amounts to direct estimation of one common expected 
	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
	$\phi_{jd}$ and/or $\chi_d$ as in model (ii). All 
	models are fit on 241 randomly selected cell lines, while performance measures 
	are estimated from the remaining 241 cell lines.

	\subsection{Results}
<<simulation_gdsc>>=
res1 <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
res2 <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
res3 <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
temp1 <- res1[, 1]
temp2 <- res2[, 1]
temp3 <- res3[, 1]
res1 <- as.matrix(res1[, -1])
res2 <- as.matrix(res2[, -1])
res3 <- as.matrix(res3[, -1])
rownames(res1) <- temp1
rownames(res2) <- temp2
rownames(res3) <- temp3
lambda <- apply(res1[rownames(res1)=="lambdaf", c(1, 2)], 2, median)
apply(res2[rownames(res2)=="lambdad", c(1, 2)], 2, median)
apply(res3[rownames(res3)=="lambdaf", c(1, 2)], 2, median)
apply(res3[rownames(res3)=="lambdad", c(1, 2)], 2, median)
est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
melbo <- apply(res[rownames(res)=="elbo", c(1, 2)], 2, median)
melbot <- apply(res[rownames(res)=="elbot", c(1, 2)], 2, median)

mphi <- median(1/res[rownames(res)=="alphaf0", 1])
mlambdaf <- median(1/res[rownames(res)=="lambdaf", 1])
@	
	Figure \ref{fig:simulations_gdsc_est1} shows the estimated
	$\bm{\alpha}_{\text{feat}}$ together with its true value for model
	(ii) in version 1 of the simulation study. Figure 
	\ref{fig:simulations_gdsc_est1}a shows that 
	$\bm{\alpha}_{\text{feat}}$ is slightly but 
	consistently underestimated. On the $\phi_{jd}$ scale this underestimation is
	almost undetectable as depicted in Figure
	\ref{fig:simulations_gdsc_est1}b. $\lambda_{\text{feat}}$ is slightly 
	overestimated with median estimate \Sexpr{round(lambda[2], 2)} (the
	true value is 1). The overestimated $\lambda_{\text{feat}}$ translates 
	into underestimated $\gamma_{jd}^2$ variances. Underestimation of variances is 
	a common phenomenon in variational Bayes approximations. In addition, 
	$\lambda_{\text{feat}}$ is a higher level scale
	parameter, which are, in general, hard to estimate. 
<<simulations_gdsc_est1, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulations with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>= 
@

	Model (i) gives a median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
	equal to the true mean of the $\phi_{jd}$, \Sexpr{round(true.phi, 2)}. 
	$\lambda_{\text{feat}}$ is slightly underestimated with median estimate
	\Sexpr{round(lambda[1], 2)} (the true value is one) and thus overestimates the 
	prior $\gamma_{jd}^2$ variance. This is likely to compensate for the common
	prior mean $\phi$, that is not able to capture part of the variation in the 
	true feature specific $\phi_{jd}$.

	The median evidence 
	lower bounds (ELBO), which approximate the marginal likelihoods of models
	(i)-(iv), are estimated as \Sexpr{round(melbo, 2)[1:3]}, and 
	\Sexpr{round(melbo, 2)[4]}, respectively. This implies that model (i),
	with estimated $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ fits
	the data best (the log Bayes factors comparing models (ii)-(iv) to model (i) 
	are \Sexpr{round(melbo[-c(1, 4)] - melbo[1], 2)}, and 
	\Sexpr{round(melbo[4] - melbo[1], 2)}, respectively, and thus highly favouring
	model (i)). The prediction MSEs for the four models on the test 
	data and gaps between training and test prediction MSE are displayed in Table
	\ref{tab:mse1}. Table \ref{tab:mse1} confirms that the models that
	include the external data (models (i) and (ii)) learn the underlying structure 
	in the data better than models (iii) and (iv) that do not include the external 
	data.
<<mse1>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp
memse <- apply(res[rownames(res)=="emse", ], 2, median)
mpmse <- apply(res[rownames(res)=="pmse", ], 2, median)
mpmset <- apply(res[rownames(res)=="pmset", ], 2, median)
tab <- rbind(memse, mpmse, mpmse - mpmset)
rownames(tab) <- c("$\\text{EMSE}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}} - \\text{PMSE}_{\\text{train}}$")
options(knitr.kable.NA="", digits=3)
kable(tab,
      col.names=colnames(tab),
      caption="Median estimation MSE, prediction MSEs, and difference between 
      prediction MSEs on the test and training data in the fixed 
      $\\tau_d^2$ simulation setting.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

	To investigate the interplay of the feature and drug variance components, we
	repeat the whole simulation study with random $\tau_d^2$. Similarly as with
	the $\gamma_{jd}^2$, we randomly create four groups of drugs and corresponding
	external drug dummy covariates. We simulate the $\tau_d^2$ as in
	(\ref{eq:tauprior}), where we set $\bm{\alpha}_{\text{drug}} =
	\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$, such
	that the $\tau_{d}^2$ of the four groups of features have prior means
	$\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The scale $\lambda_{\text{drug}}$ is
	set to one. We estimate the same four models as in the fixed $\tau_d^2$
	setting, but in addition estimate the $\tau_d^2$ priors as well. Again, we
	use two different versions: one where we estimate
	$\bm{\lambda}$ and one where we fix it to its true value.

	Figures \ref{fig:simulations_gdsc_est2} displays the estimated $\bm{\alpha}$,
	$\phi_{jd}$, and $\chi_d$ for models (i) and (ii). The scale parameters were
	estimated as $\lambda_{\text{feat}}=\Sexpr{round(rnorm(1), 2)}$ and
	$\lambda_{\text{drug}}=\Sexpr{round(rnorm(1)[1], 2)}$ in model (i). From
	Figure \ref{fig:simulations_gdsc_est2} we see that estimation of
	$\bm{\alpha}_{\text{feat}}$ is again biased. However, from
	Figure \ref{fig:simulations_gdsc_phi2} we see that the scale and ordering of
	the $\phi_{jd}$ are correclty estimated. We conjecture that the drop in
	estimation accuracy compared to the previous setting stems from the
	difficult estimation of the $\tau_d^2$ prior. Estimation of this prior
	depends on only 168 observed drugs, while the $\gamma_{jd}^2$ prior is
	is estimated from approximately $300 \cdot 168=50400$ features.
	\cite{wiel_learning_2019} show in a simple regression setting that if the
	number of samples and the number of parameters on which the EB is based are of
	the same order, empirical Bayes performs worse than when one of the two is
	large compared to the other.
	Here, the number of samples $n=241$ and $D=168$ are of the same order, leading
	to a suboptimal EB estimation performance.
<<simulations_gdsc_est2, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulations with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>= 
@

  We argue however, that even though the prior parameters are not estimated
  correctly, there is still some gain in predictive performance.
  Prediction MSEs on the test data the training-test gaps in prediction MSE are
  shown in Table \ref{tab:pmse2}. The comparisons of models (i) and
	(ii) vesus models (iii) and (iv) confirm that even though EB estimation is
	suboptimal, the external data do add something in predictive performance.

	\section{GDSC data}\label{sec:gdsc}
	\subsection{Primary data}
	The GDSC project's \cite[]{yang_genomics_2013} aim is ``to improve cancer 
	treatments by discovering therapeutic biomarkers that can be used to identify 
	patients most likely to respond to anticancer drug". Part of the project is to
	screen $>1000$ human cancer cell lines for drug sensitivities. The cell lines
	have been genetically characterised and several drug sensitivity measures are
	recorded.
	The data is freely available from \cite{garnett_systematic_2012} and consist 
	of: (i) the sensitivity measures of the cell lines 
	to the drugs, (ii) annotation of the 
	screened compounds, and (iii) the cell lines' genomic profile (mutations, 
	copy numbers, methylation profiles, and gene expression). We will attempt to 
	predict drug sensitivities of the cell lines, as quantified by half maximal
	inhibitory concentration (IC50), using the
	gene expression data. Other choices of sensitivity measures than IC50 are 
	possible, but a discussion on the pros and cons of different sensitivity
	measures is beyond the aim of this paper. A comprehensive penalized regression
	approach such as in 
	{\cite{mai_composite_2019} uses all the available molecular markers 
	instead of just gene expression (68 mutations, 426 copy numbers, 
	and 2602 gene expression levels for 498 cell lines and 97 drugs). The 
	current version of our software is not able to handle such problem sizes,
	due to the repeated calculation of posterior parameters in the empirical
	Bayes iterations. Currently, problems of up to 1000 outcomes, 1000 features,
	and 1000 samples are computationally feasible.

	We average repeated measures over cell line-drug combinations and model the
	logarithm of the IC50 values. In the following, IC50 refers 
	to the log-transformed values. 
	After iteratively removing the cell line or drug with the highest proportion 
	of missings until no more missing values remain, we end up with IC50 
	estimates for 482 cell lines on 168 drugs. As a preliminary step, we fit
	separate elastic net models for each drug with the \texttt{glmnet} package
	\cite[]{friedman_regularization_2010}, with $\alpha=0.01$ and $\lambda$
	chosen such that we select approximately 300 genes per drug.

@@

	\subsection{Results}
	To compute performance measures, we randomly split the data into training and
	test sets, containing 
	241 cell lines each, estimate the models on the training data and compute 
	performance measures on the test data. We repeated this procedure 50 times
	and average performance measures over the splits.
	The estimated 
	models are: two models like in (\ref{eq:prior}) but with different external 
	variables, named (i) NIG1 and (ii) NIG2 in the following, (iii) a model 
	as described in \cite{kpogbezan_empirical_2017}, named bSEM, (iv) a regular
	ridge model, and (v) a regular lasso model. The NIG1 incorporates no external
	variables and estimates the common prior $\gamma_{jd}^2$ and $\tau_d^2$ 
	means $\phi$ and $\chi$ directly, while the NIG2 model includes the
	developmental phase and action drug covariates, and the pathway
	feature covariate, and estimates $\bm{\alpha}$. The bSEM model is similar to 
	the NIG model in that it 
	draws the $\beta_{jd}$ from a conditionally normal distribution, but 
	fixes $\tau_d^2=1$ and draws 
	$\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of 
	from the inverse Gaussian distribution. Additionally, 
	the bSEM model can 
	only incorporate one binary external covariate that divides the features into 
	two groups, i.e., $\bm{\alpha} = \begin{bmatrix} \alpha_0 & \alpha_1 
	\end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix} 
	\tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix} 
	1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both 
	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
	estimated in a similar fashion as in Section \ref{sec:estimation}.

>>>>

	We estimate four models: (i) an NIG model that does include external 
	covariates in the estimation called NIG1 in the following, (ii) the NIG model
	estimated as in Section \ref{sec:estimation} called NIG2, and regular (iii)
	lasso and (iv) ridge models.
	Exclusion of the external covariates as in the NIG1 model
	amounts to direct estimation of one common expected 
	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
	$\phi_{jd}$ and/or $\chi_d$ as in the NIG2 model. All 
	models are fit on 241 randomly selected cell lines, while performance measures 
	are estimated from the remaining 241 cell lines.

	\subsection{Results}
<<simulation_gdsc_res1>>=
res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp
true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="elbot", ], 2, median) -
               apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="emse", ], 2, median),
             apply(res[rownames(res)=="pmse", ], 2, median),
             apply(res[rownames(res)=="pmse", ], 2, median) - 
               apply(res[rownames(res)=="pmset", ], 2, median))
tab <- round(tab, 3)
tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
                   "$\\text{ELBO}_{\\text{train}} 
                   - \\text{ELBO}_{\\text{test}}$",
                   "$\\text{EMSE}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}} 
                   - \\text{PMSE}_{\\text{train}}$")     
@	
  Figure \ref{fig:simulations_gdsc_est1} shows the estimated
	$\bm{\alpha}_{\text{feat}}$ together with its true value for NIG2 in 
	version 1 of the simulation study. Figure 
	\ref{fig:simulations_gdsc_est1}a shows that 
	$\bm{\alpha}_{\text{feat}}$ is slightly but 
	consistently underestimated. On the $\phi_{jd}$ scale this underestimation is
	almost undetectable as depicted in Figure
	\ref{fig:simulations_gdsc_est1}b. $\lambda_{\text{feat}}$ is slightly 
	overestimated with median estimate \Sexpr{round(est.lambdaf[2], 2)} (the
	true value is 1). The overestimated $\lambda_{\text{feat}}$ translates 
	into underestimated $\gamma_{jd}^2$ variances. Underestimation of variances is 
	a common phenomenon in variational Bayes approximations. In addition, 
	$\lambda_{\text{feat}}$ is a higher level scale
	parameter, which are, in general, hard to estimate. 
<<simulations_gdsc_est1, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 1 with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>= 
@

	Model NIG1 gives a median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
	equal to the true mean of the $\phi_{jd}$, \Sexpr{round(true.phi, 2)}. 
	$\lambda_{\text{feat}}$ is slightly underestimated with median estimate
	\Sexpr{round(est.lambdaf[1], 2)} (the true value is one) and thus 
	overestimates the 
	prior $\gamma_{jd}^2$ variance. This is likely to compensate for the common
	prior mean $\phi$, that is not able to capture part of the variation in the 
	true feature specific $\phi_{jd}$.

	Table \ref{tab:simulations_gdsc_tab1} displays several median performance
	measures for simulation version 1. The table includes as a reference 
	the null model, which
	is an empty model that only fits the mean of the $\y_{id}$.
	The median evidence lower bounds (ELBO) approximates the marginal
	likelihoods of models NIG1 and NIG2 on both training and test data, while 
	estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
	performance. Focussing on MSE, we see that the NIG2 model, that includes
	the external covariates, performs best, both in terms of estimation and 
	prediction. Model fit is best for the NIG1 model that does not include the
	external covariates, according to the ELBO on the training data.
	The lower MSEs confirm that the model that
	includes the external data (NIG2) learns the underlying structure 
	in the data better than model NIG1 that does not include the external 
	data.
<<simulations_gdsc_tab1>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure estimates for simulation version 1, with the best performing model in bold.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<simulation_gdsc_res2>>=
res <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp
true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
est.chi <- median(1/res[rownames(res)=="alphad0", 1])
est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="elbot", ], 2, median) -
               apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="emse", ], 2, median),
             apply(res[rownames(res)=="pmse", ], 2, median),
             apply(res[rownames(res)=="pmse", ], 2, median) - 
               apply(res[rownames(res)=="pmset", ], 2, median))
tab <- round(tab, 3)
tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
                   "$\\text{ELBO}_{\\text{train}} 
                   - \\text{ELBO}_{\\text{test}}$",
                   "$\\text{EMSE}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}} 
                   - \\text{PMSE}_{\\text{train}}$")     
@	
  Figure \ref{fig:simulations_gdsc_est2} shows the estimated
	$\bm{\alpha}_{\text{drug}}$ together with its true value for NIG2 in 
	version 2 of the simulation study. They are estimated without bias.
	Median $\lambda_{\text{drug}}=\Sexpr{round(est.lambdad[2], 2)}$ is slightly 
	overestimated compared to true value 1. 
	The median $\chi$ estimate in the NIG1 model is \Sexpr{round(est.chi, 2)},  
	which is almost equal to the true mean \Sexpr{round(true.chi, 2)}. 
	$\lambda_{\text{drug}}$ is underestimated with median estimate
	\Sexpr{round(est.lambdad[1], 2)} compared to the true value.
	Median performance measures for simulation version 2 are displayed in Table
	\ref{tab:simulations_gdsc_tab2}. Again, we see that NIG2 performs best in 
	terms of MSE. However, NIG1 performance is the same here, indicating that 
	the external covariates do not add much here.
	Model fit is again best for the NIG1 model.
<<simulations_gdsc_est2, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 2 with fixed $\\gamma_{jd}^2$.", out.width="100%", fig.asp=1/2>>= 
@

<<simulations_gdsc_tab2>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure estimates for simulation version 2, with the best performing model in bold.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<simulation_gdsc_res3>>=
res <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp
true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
est.chi <- median(1/res[rownames(res)=="alphad0", 1])
est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)

phi.ratio <- apply(apply(sapply(0:3, function(s) {
  res[rownames(res)==paste0("alphaf" ,s), colnames(res)=="NIG2"]}) %*% 
    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
      res[rownames(res)=="alphaf0", colnames(res)=="NIG2"]/m}), 2, median)
chi.ratio <- apply(apply(sapply(0:3, function(s) {
  res[rownames(res)==paste0("alphad" ,s), colnames(res)=="NIG2"]}) %*% 
    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
      res[rownames(res)=="alphad0", colnames(res)=="NIG2"]/m}), 2, median)
true.ratio <- as.numeric(1/(c(1, 1, 3, 7) %*% 
                              t(unname(model.matrix(~ factor(1:4))))))

tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="elbot", ], 2, median) -
               apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="emse", ], 2, median),
             apply(res[rownames(res)=="pmse", ], 2, median),
             apply(res[rownames(res)=="pmse", ], 2, median) - 
               apply(res[rownames(res)=="pmset", ], 2, median))
tab <- round(tab, 3)
tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
                   "$\\text{ELBO}_{\\text{train}} 
                   - \\text{ELBO}_{\\text{test}}$",
                   "$\\text{EMSE}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}} 
                   - \\text{PMSE}_{\\text{train}}$")     
@	
  In Figure \ref{fig:simulations_gdsc_est3} the 
	$\bm{\alpha}_{\text{feat}}$ and $\bm{\alpha}_{\text{drug}}$ estimated by NIG2
	are displayed together with their true values for
	simulation 3. $\bm{\alpha}_{\text{feat}}$ are underestimated, while 
	$\bm{\alpha}_{\text{drug}}$ are overestimated. These biases seem to 
	be consistent though. The median ratios $\phi_1$ to $\phi_2$, $\phi_3$, 
	$\phi_4$ and $\chi_1$ to $\chi_2$, $\chi_3$, 
	$\chi_4$ are \Sexpr{round(phi.ratio, 2)[-1]} and 
	\Sexpr{round(chi.ratio, 2)[-1]}, respectively, compared to their true values
	\Sexpr{round(true.ratio, 2)[-1]}. So in a relative sense, the estimates are
	correct. Median $\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ are
	are over- and underestimated with \Sexpr{round(est.lambdaf[2], 2)}
	and \Sexpr{round(est.lambdad[2], 2)}, respectively, compared to their true
	value 1. The NIG1 model is also consistently over- and underestimating $\phi$
	and $\chi$ with estimates \Sexpr{round(est.phi, 2)} and 
	\Sexpr{round(est.chi, 2)}, respectively (compared to the true mean 
	\Sexpr{round(true.chi, 2)}). NIG1 also over- and underestimates 
	$\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ with median estimates
	\Sexpr{round(est.lambdaf[1], 2)}
	and \Sexpr{round(est.lambdad[1], 2)}, respectively. Table 
	\ref{tab:simulations_gdsc_tab3}, with median performance measures, shows
	that NIG2 outperforms the other models in terms of prediction and estimation
	MSE.
<<simulations_gdsc_est3, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 3.", out.width="100%", fig.asp=1>>= 
@

<<simulations_gdsc_tab3>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure estimates for simulation version 3, with the best performing model in bold.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

	\section{GDSC data}\label{sec:gdsc}
	\subsection{Primary data}
	The GDSC project's \cite[]{yang_genomics_2013} aim is ``to improve cancer 
	treatments by discovering therapeutic biomarkers that can be used to identify 
	patients most likely to respond to anticancer drug". Part of the project is to
	screen $>1000$ human cancer cell lines for drug sensitivities. The cell lines
	have been genetically characterised and several drug sensitivity measures are
	recorded.
	The data is freely available from \cite{garnett_systematic_2012} and consist 
	of: (i) the sensitivity measures of the cell lines 
	to the drugs, (ii) annotation of the 
	screened compounds, and (iii) the cell lines' genomic profile (mutations, 
	copy numbers, methylation profiles, and gene expression). We will attempt to 
	predict drug sensitivities of the cell lines, as quantified by half maximal
	inhibitory concentration (IC50), using the
	gene expression data. Other choices of sensitivity measures than IC50 are 
	possible, but a discussion on the pros and cons of different sensitivity
	measures is beyond the aim of this paper. A comprehensive penalized regression
	approach such as in 
	{\cite{mai_composite_2019} uses most of the available molecular markers 
	instead of just gene expression (68 mutations, 426 copy numbers, 
	and 2602 gene expression levels for 498 cell lines and 97 drugs). The 
	current version of our software is not able to handle such problem sizes,
	due to the repeated calculation of posterior parameters in the empirical
	Bayes iterations. Currently, problems with number of outcomes in the hundreds
	and number of features and samples in the thousands are computationally 
	feasible.

	We average repeated measures over cell line-drug combinations and model the
	logarithm of the IC50 values. In the following, IC50 refers 
	to the log-transformed values. 
	After iteratively removing the cell line or drug with the highest proportion 
	of missings until no more missing values remain, we end up with IC50 
	estimates for 482 cell lines on 168 drugs.

@@

	\subsection{Results}
	As a preliminary step, we select
	features according to three strategies. The first strategy selects for each
	drug separately, the 50 features that are not in the pathway of the drug with
	the largest variance and the 50 features that are in the pathway of the drug
	with the largest variance. If less than 50 features that are in the pathway 
	are available, we select the maximum number available. This strategy ensures
	that the features are somewhat balanced with respect to the external feature
	covariate, i.e., pathway membership. The second strategy selects the 
	100 features with the largest variance, irrespective of pathway membership,
	while the last stategy selects 1000 features with the largest variance,
	again irrespective of pathway membership. These three strategies lead to
	three different analysis outcomes.

	To compute performance measures, we randomly split the data into training and
	test sets, containing 
	241 cell lines each, estimate the models on the training data and compute 
	performance measures on the test data. We repeated this procedure 50 times
	and average performance measures over the splits.
	The estimated 
	models are: four models like in (\ref{eq:prior}) but with different external 
	variables, named NIG1, NIG2, NIG3, and NIG4 in the following. NIG1 estimates
	only one overall $\phi$ and one overall $\chi$, NIG2 uses the external 
	covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
	$\bm{\alpha}_{\text{drug}}$, while NIG3 and NIG4 estimate only 
	$\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
	fixed. In addition, we estimate regular lasso and ridge models.

	% The bSEM model is similar to 
	% the NIG model in that it 
	% draws the $\beta_{jd}$ from a conditionally normal distribution, but 
	% fixes $\tau_d^2=1$ and draws 
	% $\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of 
	% from the inverse Gaussian distribution. Additionally, 
	% the bSEM model can 
	% only incorporate one binary external covariate that divides the features into 
	% two groups, i.e., $\bm{\alpha} = \begin{bmatrix} \alpha_0 & \alpha_1 
	% \end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix} 
	% \tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix} 
	% 1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both 
	% $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
	% estimated in a similar fashion as in Section \ref{sec:estimation}.

<<analysis_gdsc_fit1>>=
fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
temp <- fit[, 1]
fit <- as.matrix(fit[, -1])
rownames(fit) <- temp

res <- read.table("results/analysis_gdsc_res1.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp

est1 <- cbind(fit[, 1], c(NA, 0, 0, NA), fit[, 2])
colnames(est1) <- c("intercept", "not in pathway", "in pathway")
est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0), fit[, c(7:6)], c(NA, 0, NA, 0))
colnames(est2) <- c("intercept", "experimental", "development", "approved",
                    "unkown", "targeted", "cytotoxic")
